Tail Indices and Scale Parameters in Financial Time Series

### 2017-03-29 papers/FX/draft.tex
Revised the scripts.
Added an introductory paragraph at the beginning of the section of empirical results.

\section{Introduction}
It is generally believed that the utility function of an
investor is increasing and concave -- as John H. Cochrane
put it, ``the last bite is never as satisfying as the first''
\cite{cochrane2009asset}. This implies investors are considerably
more concerned about large losses than are cheered by large gains.
Furthermore, it is a stylized fact that equity return distributions
are heavy-tailed (see e.g. Jansen and de Vries \cite{jansen1991frequency}).
In particular, the lower-tail of an equity distribution is often heavier
than the upper-tail of the same distribution.

We show in this paper that concerns of the down-side risk and the
heavy-tail property of equity returns  imply that the lower-tail indices
of actively traded equities in a rational market must be fairly close to
one another.

%% A cornerstone of this conclusion is the empirical fact that equity
%% return distributions are heavy-tailed (see e.g. Jansen and de Vries
%% (1991) \cite{jansen1991frequency}). If investors and speculators alike
%% display concern for downside risk, as formulated via a utility function,
%% it is important to model this risk adequately.

In section \ref{sec:GDA}, we explicitly recognize the behavioral
concern for downside risk in the investor's evaluation of portfolios
using the framework of Generalized Disappointment Aversion (GDA)
introduced by Routledge and Zin \cite{routledge2010generalized}. GDA
is an extension of the concept of Disappointment Aversion (DA) of Gul
(1991) \cite{gul1991theory}. The Gul paper derives the DA from first
principles (axiomatic).

@@

\section{Empirical Results}
\subsection{Hill estimates of lower tail indices}
\label{sec:Hill}
There have been several estimators for the tail index proposed in the
literature, e.g. Hill (1975) \cite{hill1975simple}, Pikands (1975)
\cite{pickands1975statistical}, Dekkers and Einmahl and De Haan (1989)
\cite{dekkers1989moment}. See also Embrechts and Kl\"uppelberg and
Mikosch (1997)\cite{Embrechts1997}. The most popular of these is
arguably Hill (1975) \cite{hill1975simple}:
\[
\hat \alpha = \left[
  {1 \over k} \sum_{i=1}^k \ln {x_{(i)} \over x_{(k+1)}}
  \right]^{-1}
\]
where $x_{(i)}$ is the $i$-th upper order statistic of a sample
$x_1, x_2, ..., x_n$. It has been shown that the Hill estimator is
applicable to 
\begin{itemize}
\item iid random variables (Mason (1982) \cite{mason1982laws})
\item a weakly dependent sequence (Rootz{\'e}n, Leadbetter,
  and de Haan \cite{rootzen1992tail})
\item a linear process (Resnick and St{\u{a}}ric{\u{a}} (1995)
  \cite{resnick1995consistency}).
\end{itemize}
In these cases it is shown that, if $k \to \infty$ and $k/n \to 0$
as $n \to \infty$, the Hill estimator is weakly consistent, i.e.
\[
\sqrt k (\hat \alpha - \alpha) \overset{d}{\to} N(0, \alpha^2)
\]

Figure \ref{fig:Energy_lower},
\ref{fig:Consumer_Staples_lower},
and \ref{fig:Information_Technology_lower}
show the Hill estimates of the
lower tail indices of the ``energy'', ``consumer staples'' and ``information
technology'' sectors of the S\&P 500 index, respectively.
\begin{figure}[htb!]
  \centering
  \begin{subfigure}[b]{\linewidth}
    \includegraphics[width=\textwidth, trim={0, 0.8cm, 0, 2cm}, clip]
                    {Energy_lower.pdf}
    \caption{``Energy'' sector of S\&P 500}
    \label{fig:Energy_lower}
  \end{subfigure}
  \begin{subfigure}[b]{\linewidth}
    \includegraphics[width=\textwidth, trim={0, 0.8cm, 0, 2cm}, clip]
                    {Consumer_Staples_lower.pdf}
    \caption{``Consumer Staples'' sector of S\&P 500}
    \label{fig:Consumer_Staples_lower}
  \end{subfigure}
  \begin{subfigure}[b]{\linewidth}
    \includegraphics[width=\textwidth, trim={0, 0.8cm, 0, 2cm}, clip]
                    {Information_Technology_lower.pdf}
    \caption{``Information Technology'' sector of S\&P 500}
    \label{fig:Information_Technology_lower}
  \end{subfigure}
  \caption{\small Hill estimates of lower tail indices of
    daily return series in sectors of Standard \& Poor
    500 index. Each circle coresponds to a Hill estimate; the grey
    triangles above and below it mark the 97.5\% and 2.5\% quantiles
    of its approximate normal distribution.
    The low and high red lines mark the median of the 2.5\% quantiles
    and median of the 97.5\% quantiles, respectively.
    The data are from {\it Yahoo Finance} and the labels on
    the horizontal axis are Yahoo symbols of the stocks. The data span
    from 2010-01-01 to 2015-01-01 and comprise $n=1304$
    observations. $k=50$ lower order statistics are used
    for computing $\alpha_n$. 
  }
\end{figure}
One can see that all the confidence intervals of stocks in the
``Energy'' and ``Consumer Staples'' sectors, as well as those of a
large portion of stocks in the ``Information Technology'' sector,
overlap. This naturally prompts an investigation into the extreme
quantile estimates of the same stocks.

@@

\subsection{Hoga's Test}
Hoga \cite{Hoga2017} proposed his test for a time series
$\{X_i\}_{i=1}^n$ with tail functions
$\bar F_i(\cdot)$, i.e. $\P(X_i > x) = \bar F_i(x)$. The null
hypothesis is
\begin{hypothesis}
  $\mathcal H_0$: For a given $p$, $\exists t^* \in [t_0, 1 - t_0]$
  such that
  \[
  \bar F_1^{-1}(p) = \bar F_2^{-1}(p) = \cdots =
  \bar F_{\floor{t^* n}}^{-1}(p) \neq
  \bar F_{\ceil{t^* n}}^{-1}(p) = \cdots = F_n^{-1}(p)
  \]
\end{hypothesis}
where $t_0 \in (0, 1/2)$. The alternative hypothesis is that no such
$t^*$ exists. Hoga proposed the following test statistics:
\[
Q_1 = \sup_{t \in [t_0, 1 - t_0]}
\left\{
  {
    \left[
      t (1 - t) \ln\left(
      \hat x_p(0, t)
      \over
      \hat x_p(t, 1)
      \right)
    \right]^2
    \over
    \int_{t_0}^t
    \left[
      s \ln \left(
      \hat x_p(0, s)
      \over
      \hat x_p(0, t)
      \right)
    \right]^2 ds
    +
    \int_{t}^{1 - t_0}
    \left[
      (1 - s) \ln \left(
      \hat x_p(s, 1)
      \over
      \hat x_p(t, 1)
      \right)
    \right]^2 ds
  }
\right\}
\]
Under the null hypothesis, Hoga showed that as $n \to \infty$,
\begin{equation}
  \label{eq:Q1_distr}
  Q_1 \overset{d}{\to}
  \sup_{t \in [t_0, 1 - t_0]} q_t
\end{equation}
where
\[
q_t = 
  {
    [W(t) - t W(1)]^2
    \over
    \int_{t_0}^t [W(s) - {s \over t}W(t)]^2 ds
    +
    \int_{t}^{1 - t_0}
    \left[
      W(1) - W(s) - {1 - s \over 1 - t}
      (W(1) - W(t))
      \right]^2 ds
  }
\]
$W(t)$ with $t \in (0, 1)$ is a standard Brownian motion.

@@

To test whether two series share the same tail index, we concatenate
the two series and apply the Hoga test. For the ``Energy'' and the
``Consumer Staples'' sector, we summarise the results in figure
\ref{fig:Hoga_Energy_pair} and \ref{fig:Hoga_consumer_Staples_pair}.
\begin{figure}[htb!]
  \begin{subfigure}[b]{0.5\linewidth}
    \includegraphics[width=\textwidth]{Hoga_Energy_pair.pdf}
    \caption{``Energy''}
    \label{fig:Hoga_Energy_pair}
  \end{subfigure}
  \begin{subfigure}[b]{0.5\linewidth}
    \includegraphics[width=\textwidth]{Hoga_Consumer_Staples_pair.pdf}
    \caption{``Consumer Staples''}
    \label{fig:Hoga_consumer_Staples_pair}
  \end{subfigure}
  \caption{Hoga Test Statistics of stocks in the ``Energy'' and
    ``Consumer Staples'' sectors of
    S\&P 500. The green, blue and red colors represent, respectively,
    exceedances of the 85\%, 90\% and 95\% quantiles of the distribution of
    \eqref{eq:Q1_distr}.
    Grey points stand for pairs for which the test statistic is below
    the 85\% quantile;
    black points represent pairs for which
    computation of the test statistics fails for given precision
    requirements and time limits}.
\end{figure}
Clearly, these figures show that the null
hypothesis of an equal tail index is rejected for more pairs in the
``Energy'' sector than it is for those in the ``Consumer Staples''
sector. This suggests that lower tail indices of stocks in the
``energy'' sector are more spread out than are those of the ``Consumer
Staples'' sector.

@@

Figure \ref{fig:Energy_alpha_K},
\ref{fig:Consumer_Staples_alpha_K},
and \ref{fig:Information_Technology_alpha_K}
show the estimates of $K$ alongside the Hill estimates of $\alpha$ of
the ``energy'', ``consumer staples'' and ``information
technology'' sectors of the S\&P 500 index, respectively.
\begin{figure}[htb!]
  \centering
  \begin{subfigure}[b]{0.45\linewidth}
    \includegraphics[width=\textwidth]
                    {Energy_alpha_K.pdf}
                    \caption{``Energy''}
                    \label{fig:Energy_alpha_K}
  \end{subfigure}
  \begin{subfigure}[b]{0.45\linewidth}
    \includegraphics[width=\textwidth]
                    {Energy_alpha_K_ci.pdf}
                    \caption{``Energy''}
                    \label{fig:Energy_alpha_K_ci}
  \end{subfigure}
  \begin{subfigure}[b]{0.45\linewidth}
    \includegraphics[width=\textwidth]
                    {Consumer_Staples_alpha_K.pdf}
                    \caption{``Consumer Staples''}
                    \label{fig:Consumer_Staples_alpha_K}
  \end{subfigure}
  \begin{subfigure}[b]{0.45\linewidth}
    \includegraphics[width=\textwidth]
                    {Consumer_Staples_alpha_K_ci.pdf}
                    \caption{``Information Technology''}
                    \label{fig:Consumer_Staples_alpha_K_ci}
  \end{subfigure}
  \begin{subfigure}[b]{0.45\linewidth}
    \includegraphics[width=\textwidth]
                    {Information_Technology_alpha_K.pdf}
                    \caption{``Information Technology''}
                    \label{fig:Information_Technology_alpha_K}
  \end{subfigure}
  \begin{subfigure}[b]{0.45\linewidth}
    \includegraphics[width=\textwidth]
                    {Information_Technology_alpha_K_ci.pdf}
                    \caption{``Information Technology''}
                    \label{fig:Information_Technology_alpha_K_ci}
  \end{subfigure}
  \caption{\small Maximum likelihood estimates of $K$, the scale
    parameter, of stocks in sectors of the S\&P 500 index. The points
    are the
    estimated values; the red bars are the upper bound of the 98\%
    confidence interval. The lower bound is 0.
  }
\end{figure}

>>>>

\section{Introduction}
It is generally believed that the utility function of an
investor is increasing and concave -- as John H. Cochrane
put it, ``the last bite is never as satisfying as the first''
\cite{cochrane2009asset}. This implies investors are repelled to a
larger extent by large losses than they are tempted by large gains. 
Furthermore, it is a stylized fact that equity return distributions
are heavy-tailed (see e.g. Jansen and de Vries \cite{jansen1991frequency}).
In particular, the lower-tail of an equity distribution is often heavier
than the upper-tail of the same distribution.

we explore in this paper the market implications of investor
preferences and the heavy-tail property of equity return
distributions. The rest of this paper is organized into the following
sections:
In section \ref{sec:empirical_results} we present a survey of
statistical characteristics of equity returns. Then 
in section \ref{sec:GDA}, we explicitly recognize the behavioral
concern for downside risk in the investor's evaluation of portfolios
using the framework of Generalized Disappointment Aversion (GDA)
introduced by Routledge and Zin \cite{routledge2010generalized}. GDA
is an extension of the concept of Disappointment Aversion (DA) of Gul
(1991) \cite{gul1991theory}. The Gul paper derives the DA from first
principles (axiomatic).
In this section we infer characteristics of equity returns from the
GDA theory and compare them with the empirical results.

@@

\section{Empirical Results}
\label{sec:empirical_results}
As has been mentioned, we are interested in the heavy-tail properties
of equity returns. Let $X_t$ be an equity return series. If the
distribution of $X_t$ is heavy-tailed, we can write
\begin{eqnarray*}
  \lim_{x \to \infty} {x^{\beta}\P(X_t > x) \over L_+(x)} &=& 1 \\
  \lim_{x \to \infty} {x^{\alpha}\P(X_t < -x) \over L_-(x)} &=& 1
\end{eqnarray*}
where $L_+(x)$ and $L_-(x)$ are slowly varying functions satisfying
\[
\lim_{x \to \infty} {L_*(a x) \over L_*(x)} = 1
\]
for all $a > 0$. Here ``*'' can be either ``+'' or ``-''. In the
simplest case, $L_+(x)$ and $L_(x)$ are constants and the distribution
has power-law tails. Let $J_*(1/x) = L_*(x)$. A Taylor expansion at
$x=\infty$ gives
\[
J_*(1/x) = J_*(0) + {J'_*(0) \over x} + O(x^{-2})
\]
Thus
\[
x^{-\alpha} L_-(x) = J_-(0) x^{-\alpha} + J'_-(0) x^{-\alpha - 1} + O(x^{-\alpha-2})
\]
when $x \to \infty$. Clearly the tail indices $\alpha$ and $\beta$ as
well as the scale parameters $J_-(0)$ and $J_+(0)$ play crutial roles
in determining the tail distribution and hence in asset allocation and
risk management. These parameters of the equities in a market are
characteristics of the market and provide a useful description of the
market. One can also use them to help build a model of the equities in
the market. With these motives in mind, we present a survey of the
tail indices and scale parameters of 3 sectors, namely ``Energy'',
``Consumer Staples'' and ``Information Technology'', of the S\&P 500
index.

\subsection{Hill estimates of lower tail indices}
\label{sec:Hill}
There have been several estimators for the tail index proposed in the
literature, e.g. Hill (1975) \cite{hill1975simple}, Pikands (1975)
\cite{pickands1975statistical}, Dekkers and Einmahl and De Haan (1989)
\cite{dekkers1989moment}. See also Embrechts and Kl\"uppelberg and
Mikosch (1997)\cite{Embrechts1997}. The most popular of these is
arguably Hill (1975) \cite{hill1975simple}:
\[
\hat \alpha = \left[
  {1 \over k} \sum_{i=1}^k \ln {x_{(i)} \over x_{(k+1)}}
  \right]^{-1}
\]
where $x_{(i)}$ is the $i$-th upper order statistic of a sample
$x_1, x_2, ..., x_n$. It has been shown that the Hill estimator is
applicable to 
\begin{itemize}
\item iid random variables (Mason (1982) \cite{mason1982laws})
\item a weakly dependent sequence (Rootz{\'e}n, Leadbetter,
  and de Haan \cite{rootzen1992tail})
\item a linear process (Resnick and St{\u{a}}ric{\u{a}} (1995)
  \cite{resnick1995consistency}).
\end{itemize}
In these cases it is shown that, if $k \to \infty$ and $k/n \to 0$
as $n \to \infty$, the Hill estimator is weakly consistent, i.e.
\[
\sqrt k (\hat \alpha - \alpha) \overset{d}{\to} N(0, \alpha^2)
\]

Figure \ref{fig:Energy_lower},
\ref{fig:Consumer_Staples_lower},
and \ref{fig:Information_Technology_lower}
show the Hill estimates of the
lower tail indices of the ``energy'', ``consumer staples'' and ``information
technology'' sectors of the S\&P 500 index, respectively.
\begin{figure}[htb!]
  \centering
  \begin{subfigure}[b]{\linewidth}
    \includegraphics[width=\textwidth, trim={0, 0.8cm, 0, 2cm}, clip]
                    {Energy_lower.pdf}
    \caption{``Energy'' sector of S\&P 500}
    \label{fig:Energy_lower}
  \end{subfigure}
  \begin{subfigure}[b]{\linewidth}
    \includegraphics[width=\textwidth, trim={0, 0.8cm, 0, 2cm}, clip]
                    {Consumer_Staples_lower.pdf}
    \caption{``Consumer Staples'' sector of S\&P 500}
    \label{fig:Consumer_Staples_lower}
  \end{subfigure}
  \begin{subfigure}[b]{\linewidth}
    \includegraphics[width=\textwidth, trim={0, 0.8cm, 0, 2cm}, clip]
                    {Information_Technology_lower.pdf}
    \caption{``Information Technology'' sector of S\&P 500}
    \label{fig:Information_Technology_lower}
  \end{subfigure}
  \caption{\small Hill estimates of lower tail indices of
    daily return series in sectors of Standard \& Poor
    500 index. Each circle coresponds to a Hill estimate; the grey
    triangles above and below it mark the 97.5\% and 2.5\% quantiles
    of its approximate normal distribution.
    The low and high red lines mark the median of the 2.5\% quantiles
    and median of the 97.5\% quantiles, respectively.
    The data are from {\it Yahoo Finance} and the labels on
    the horizontal axis are Yahoo symbols of the stocks. The data span
    from 2010-01-01 to 2015-01-01 and comprise $n=1304$
    observations. $k=50$ lower order statistics are used
    for computing $\alpha_n$. 
  }
\end{figure}
One can see that all the confidence intervals of stocks in the
``Energy'' and ``Consumer Staples'' sectors, as well as those of a
large portion of stocks in the ``Information Technology'' sector,
overlap. This suggests that the stocks in each sector may have very
similar tail indices. Hoga's test on the change of extreme quantiles
may provide some insight about how similar these tail indices are,
because different tail indices are likely to result in different
extreme quantiles. Nevertheless, changes in the extreme quantiles can
be the result of changed tail indices but can also be the result of
changed scale parameters. Thus we first investigate the scale
parameters of stocks in the same sectors before applying Hoga's test.

\subsection{Hill estimates of lower-tail scale parameters}

@@

\subsection{Hoga's Test}
\label{sec:Hoga}

Hoga \cite{Hoga2017} proposed his test for a time series
$\{X_i\}_{i=1}^n$ with tail functions
$\bar F_i(\cdot)$, i.e. $\P(X_i > x) = \bar F_i(x)$. The null
hypothesis is
\begin{hypothesis}
  $\mathcal H_0$: For a given $p$, $\exists t^* \in [t_0, 1 - t_0]$
  such that
  \[
  \bar F_1^{-1}(p) = \bar F_2^{-1}(p) = \cdots =
  \bar F_{\floor{t^* n}}^{-1}(p) \neq
  \bar F_{\ceil{t^* n}}^{-1}(p) = \cdots = F_n^{-1}(p)
  \]
\end{hypothesis}
where $t_0 \in (0, 1/2)$. The alternative hypothesis is that no such
$t^*$ exists. Hoga proposed the following test statistics:
\[
Q_1 = \sup_{t \in [t_0, 1 - t_0]}
\left\{
  {
    \left[
      t (1 - t) \ln\left(
      \hat x_p(0, t)
      \over
      \hat x_p(t, 1)
      \right)
    \right]^2
    \over
    \int_{t_0}^t
    \left[
      s \ln \left(
      \hat x_p(0, s)
      \over
      \hat x_p(0, t)
      \right)
    \right]^2 ds
    +
    \int_{t}^{1 - t_0}
    \left[
      (1 - s) \ln \left(
      \hat x_p(s, 1)
      \over
      \hat x_p(t, 1)
      \right)
    \right]^2 ds
  }
\right\}
\]
Under the null hypothesis, Hoga showed that as $n \to \infty$,
\begin{equation}
  \label{eq:Q1_distr}
  Q_1 \overset{d}{\to}
  \sup_{t \in [t_0, 1 - t_0]} q_t
\end{equation}
where
\[
q_t = 
  {
    [W(t) - t W(1)]^2
    \over
    \int_{t_0}^t [W(s) - {s \over t}W(t)]^2 ds
    +
    \int_{t}^{1 - t_0}
    \left[
      W(1) - W(s) - {1 - s \over 1 - t}
      (W(1) - W(t))
      \right]^2 ds
  }
\]
$W(t)$ with $t \in (0, 1)$ is a standard Brownian motion.

@@

To test whether two series share the same tail index, we concatenate
the two series and apply the Hoga test. For the ``Energy'' and the
``Consumer Staples'' sector, we summarise the results in figure
\ref{fig:Hoga_Energy_pair} and \ref{fig:Hoga_consumer_Staples_pair}.
\begin{figure}[htb!]
  \begin{subfigure}[b]{0.5\linewidth}
    \includegraphics[width=\textwidth]{Hoga_Energy_pair.pdf}
    \caption{``Energy''}
    \label{fig:Hoga_Energy_pair}
  \end{subfigure}
  \begin{subfigure}[b]{0.5\linewidth}
    \includegraphics[width=\textwidth]{Hoga_Consumer_Staples_pair.pdf}
    \caption{``Consumer Staples''}
    \label{fig:Hoga_consumer_Staples_pair}
  \end{subfigure}
  \caption{Hoga Test Statistics of stocks in the ``Energy'' and
    ``Consumer Staples'' sectors of
    S\&P 500. The green, blue and red colors represent, respectively,
    exceedances of the 85\%, 90\% and 95\% quantiles of the distribution of
    \eqref{eq:Q1_distr}.
    Grey points stand for pairs for which the test statistic is below
    the 85\% quantile;
    black points represent pairs for which
    computation of the test statistics fails for given precision
    requirements and time limits.}
\end{figure}
Clearly, these figures show that the null
hypothesis of an equal tail index is rejected for more pairs in the
``Energy'' sector than it is for those in the ``Consumer Staples''
sector. This suggests that lower tail indices of stocks in the
``energy'' sector are more spread out than are those of the ``Consumer
Staples'' sector.

@@

Figure \ref{fig:Energy_alpha_K},
\ref{fig:Consumer_Staples_alpha_K},
and \ref{fig:Information_Technology_alpha_K}
show the estimates of $K$ alongside the Hill estimates of $\alpha$ of
the ``energy'', ``consumer staples'' and ``information
technology'' sectors of the S\&P 500 index, respectively.
\begin{figure}[htb!]
  \centering
  \begin{subfigure}[b]{0.45\linewidth}
    \includegraphics[width=\textwidth]
                    {Energy_alpha_K.pdf}
                    \caption{``Energy''}
                    \label{fig:Energy_alpha_K}
  \end{subfigure}
  \begin{subfigure}[b]{0.45\linewidth}
    \includegraphics[width=\textwidth]
                    {Energy_alpha_K_ci.pdf}
                    \caption{``Energy''}
                    \label{fig:Energy_alpha_K_ci}
  \end{subfigure}
  \begin{subfigure}[b]{0.45\linewidth}
    \includegraphics[width=\textwidth]
                    {Consumer_Staples_alpha_K.pdf}
                    \caption{``Consumer Staples''}
                    \label{fig:Consumer_Staples_alpha_K}
  \end{subfigure}
  \begin{subfigure}[b]{0.45\linewidth}
    \includegraphics[width=\textwidth]
                    {Consumer_Staples_alpha_K_ci.pdf}
                    \caption{``Consumer Staples''}
                    \label{fig:Consumer_Staples_alpha_K_ci}
  \end{subfigure}
  \begin{subfigure}[b]{0.45\linewidth}
    \includegraphics[width=\textwidth]
                    {Information_Technology_alpha_K.pdf}
                    \caption{``Information Technology''}
                    \label{fig:Information_Technology_alpha_K}
  \end{subfigure}
  \begin{subfigure}[b]{0.45\linewidth}
    \includegraphics[width=\textwidth]
                    {Information_Technology_alpha_K_ci.pdf}
                    \caption{``Information Technology''}
                    \label{fig:Information_Technology_alpha_K_ci}
  \end{subfigure}
  \caption{\small Maximum likelihood estimates of $K$, the scale
    parameter, of stocks in sectors of the S\&P 500 index. The points
    are the
    estimated values; the bars are the upper bound of the 98\%
    confidence interval. The lower bound is 0.
  }
\end{figure}
